
```{r define-dirs, include = F}
DATA_DIR <- file.path("data")
HW_NAME <- "hw07_data-manipulation-tidyr-dplyr-join-ggplot-ddply-spin"
HW_DIR <- file.path("hw", HW_NAME)

```{r spin-code, eval = F, include = F}
# Note for clarity:
# Instead of manually creating both a .R and a .Rmd files that have almost
# identical R code, this .R script is the only file I am writing manually.
# I am using the `knitr::spin` function to turn this R script into Rmarkdown
# and then converting it into markdown and HTML formats in the code. This code
# is wrapped around an if(FALSE) so that it is not executed when the script is ran,
# and it will only do this R -> markdown/HTML conversion when these lines are
# ran explicitly. 
library(knitr)
library(markdown)

spinMyR <- function() {
	# - anything R related uses knit_root.dir (such as reading files)
	# - anything knitr related uses knit_base.dir
	# - fig.path is set RELATIVE to knit_base.dir, BUT because of a bug the folder
	#   also gets created relative to the root.dir. If all the variables aren't
	#   set up properly, figures won't show
	opts_knit$set(root.dir = getwd())
	opts_knit$set(base.dir = HW_DIR)
	opts_chunk$set(fig.path = file.path("markdown-figs", .Platform$file.sep))
	opts_chunk$set(tidy = FALSE)
	dir.create(
		file.path(opts_knit$get("base.dir"), opts_chunk$get("fig.path")),
		showWarnings = FALSE)
	
	spin(file.path(HW_DIR, paste0(HW_NAME, ".R")), knit = F)
	knit(file.path(HW_DIR, paste0(HW_NAME, ".Rmd")),
			 file.path(HW_DIR, paste0(HW_NAME, ".md")))
	markdownToHTML(file.path(HW_DIR, paste0(HW_NAME, ".md")),
								 file.path(HW_DIR, paste0(HW_NAME, ".html")))
	# because of a bug in knitr/markdown, the figs directory is created in the
	# working directory as well but with nothing in it
	suppressWarnings(unlink("markdown-figs", recursive = TRUE))
}
if (FALSE) {
	spinMyR()
}
```

# Homework 7 - Data wrangling and manipulation
Dean Attali  
Oct 2014 

Last updated: `r Sys.time()`

## Overview
In this assignment, I found a dataset with information about cancer incidences
and used various libraries (`plyr`, `dplyr`, `tidyr`, `ggplot2`) to manipulate
the data and perform some basic data exploration

## Datasets
I downloaded two datasets from [the United States CDC](http://wonder.cdc.gov/cancer.html)
that describe cancer occurrences in the US between 1999 to 2011. One dataset
holds the number of cancer cases per cancer type, while the other dataset
holds the number of deaths per cancer type.  

I also downloaded data from [The World Bank](http://data.worldbank.org/)
that provides the population of every country in every year over the past several
decades.  I only used the US data from it to know what the total US population
was at every year that we have cancer data for.

## Getting down to business
Less talkin', more codin'!

```{r load-libs, include = F}
library(plyr)
library(dplyr)
library(tidyr)
library(ggplot2)

```{r define-tolowerfirst, include = F}
tolowerfirst <- function(x) {
	return(paste0(tolower(substring(x, 1, 1)), substring(x, 2)))
}
```

### Read and clean the datasets
#### Dataset 1 - Number of cancer deaths

```{r data-deaths-read}
```

Read in the dataset that contains the number of deaths per cancer type per year
(I convert the dataframe to a tbl_df just for better visualization purposes)

```{r }
(deathsDat <- tbl_df(
	read.table(file.path(DATA_DIR, "cancerDeathsUS.txt"), header = T)))
(levels(deathsDat$Leading.Cancer.Sites))
(unique(deathsDat$Year))
```

We can see the dataset has `r nrow(deathsDat)` observations and `r ncol(deathsDat)`
variables.  There are `r length(levels(deathsDat$Leading.Cancer.Sites))` different
cancer sites for the years `r min(deathsDat$Year)` to `r max(deathsDat$Year)`.

```{r data-deaths-clean}
```

I want to do a little bit of cleaning:  
- change the column names to begin with a lower-case letter (personal preference)
(I'm using a function that I wrote, look at the source of the script to see it)  
- drop the two columns that represent "codes", they aren't needed
- rename the column that has periods in its name (personal preference)
- remove the "All Sites Combined" level

```{r }
dDatClean <- deathsDat
colnames(dDatClean) <- tolowerfirst(colnames(dDatClean))
dDatClean <- dDatClean %>%
	dplyr::select(leading.Cancer.Sites, year, deaths) %>%
	rename(cancerLocation = leading.Cancer.Sites)
dDatClean <- dDatClean %>%
	filter(cancerLocation != "All Sites Combined") %>%
	droplevels
print(dDatClean)
```

Looks good!
#### Dataset 2 - Number of cancer cases

```{r data-cases-read}
```

Read in the dataset that contains the number of incidences per cancer type per year
I will perform the same basic cleaning as on the previous dataset, since they
both came from the same source and have the same structure

```{r }
casesDat <- tbl_df(
	read.table(file.path(DATA_DIR, "cancerIncidenceUS.txt"), header = T))
cDatClean <- casesDat
colnames(cDatClean) <- tolowerfirst(colnames(cDatClean))
cDatClean <- cDatClean %>%
	dplyr::select(leading.Cancer.Sites, year, count) %>%
	rename(cancerLocation = leading.Cancer.Sites,
				 cases = count)
cDatClean <- cDatClean %>%
	filter(cancerLocation != "All Sites Combined") %>%
	droplevels
print(cDatClean)
print(levels(cDatClean$cancerLocation))
```

Looks good! This dataset has the exact same dimensions as the deaths dataset,
which is what we expected. 
### Get the two datasets ready to be merged
Next I'd like to take these two datasets and merge them together.  Specifically,
for every combination of cancer location + year, I want to merge the two datasets
so that I will have the number of cases and deaths in the same dataframe.
In order for this to work, we need to make sure that the two sources have exactly
the same levels for the cancer location and year variables. But there is a small
problem with the current data that you might have noticed - one of the cancer
locations is represented with a different name in the two datasets.  
To demonstrate this, here is the set difference between the union of the cancer
locations in both datasets and the intersection of them (ie. this shows
which cancer locations are not shared by the two)

```{r show-diff-levels}
setdiff(
	union(levels(dDatClean$cancerLocation),
				levels(cDatClean$cancerLocation)),
	intersect(levels(dDatClean$cancerLocation),
						levels(cDatClean$cancerLocation)))

```{r fix-diff-levels}
```

There is an easy fix: just change the name of that level in one of the datasets
to match the other

```{r }
identical(levels(cDatClean$cancerLocation), levels(dDatClean$cancerLocation))
cDatClean$cancerLocation <-
	cDatClean$cancerLocation %>%
	revalue(c("Urinary Bladder, invasive and in situ" = "Urinary Bladder"))
identical(levels(cDatClean$cancerLocation), levels(dDatClean$cancerLocation))
```

Now we have proof that the levels are identical
### Do the Join (aka merge)
Now the two datasets are ready to be merged together. There are many R ways
to do this, but I will use the `dplyr::left_join` approach.

```{r merge}
cancerData <- left_join(dDatClean, cDatClean, by = c("cancerLocation", "year"))
print(cancerData)
```

Success! That was easy.
### Tidying the data
Now that we have a dataset with the number of cases and deaths of every major
cancer in the US per year, we need to get the data into a tidy form so that it
will be easier to do compuations/visualization on it.  Right now the data is in
a fat/wide format, and we want to get it to a long/tall format.  Another thing
I want to do is to sort the observations by year instead of by cancer type.

```{r tidy-cancer-data}
cancerData <- cancerData %>%
	gather(stat, freq, deaths, cases) %>%
	arrange(year, cancerLocation)
print(cancerData)
```

Hooray for `tidyr`! Also, doesn't this row ordering make you happier?
(Perhaps for you it doesn't....?)
### Plot the data and save figure to file
I'd like to plot the number of cases/deaths for every cancer type in each year
as a line graph, with each cancer type being a line.

#### Re-order cancer type levels
In order to get the legend in the plot to nicely match up with the order
of the data in the plot, we need to rearrange the order of the cancer type
factor (currently it's alphabetical). To do this, I first construct a vector
holding the order of the cancer types by which one had the most cases in the
most recent timepoint, and then I recreate the factor using this ordering. 

```{r reorder-cancer-location}
cancerLocationsOrder <- cancerData %>%
	filter(stat == "cases",
				 year == max(year)) %>%
	arrange(desc(freq)) %>%
	first %>%
	as.character
cancerData$cancerLocation <-
	factor(cancerData$cancerLocation, levels = cancerLocationsOrder)
```

For brevity, I am not printing out the new order, but the plot will give
us confirmation that it worked
### Plot (providing 22 custom colours)
I couldn't find a pre-defined colour palette that I was satisfied with that
had so many colours, so I created one. Usually when dealing with less levels,
it's better to use an existing palette, such as from `RColorBrewer`.  
After plotting, I also save the plot as a PDF

```{r plot-basic-data, fig.width = 11, fig.align = "center"}
c22 <- c("dodgerblue2","#E31A1C", # red
				 "green4",
				 "#6A3D9A", # purple
				 "#FF7F00", # orange
				 "black","gold1",
				 "skyblue2","#FB9A99", # lt pink
				 "palegreen2",
				 "#CAB2D6", # lt purple
				 "#FDBF6F", # lt orange
				 "gray70", "khaki2", "maroon", "orchid1", "deeppink1", "blue1",
				 "darkturquoise", "green1", "yellow4", "brown")
p <-
	ggplot(cancerData, aes(x = year, y = freq)) +
	geom_point(aes(col = cancerLocation, group = cancerLocation), size = 2) +
	geom_line(aes(col = cancerLocation, group = cancerLocation), size = 0.7) +
	facet_wrap(~ stat) +
	theme_bw(15) +
	scale_colour_manual(values = c22)
print(p)
ggsave(file.path(HW_DIR, "cancerTypesStats.pdf"), p)
```

This is perhaps not the bestest way to show this data, but it suffices
as a basic way to see the data.  It's interesting how lung and bronchus cancers
have so many more deaths than any other cancer, even though there are two other
cancers with similar incidences. I will not make any more comments about the
plot since the purpose of this assignment is more about data manipulation than plotting
### Practice with ddply, lapply, dlply, batch reading of files
Just as a fun exercise, now I will take the cancer data, split it up by
cancer location, and write the output of every cancer type to a separate file.
Then I will read all the files and concatenate them together to recreate the
original data.
First use `ddply` to write a file for every cancer type

```{r ddply-write-files}
invisible( # invisible = I don't want to see the output from ddply
	ddply(cancerData, ~cancerLocation,
				function(x) {
					fileName <- file.path(
						HW_DIR,
						paste0("cDatTest-", gsub(" ", "_", x$cancerLocation[1]), ".csv"))
					write.table(dplyr::select(x, -cancerLocation),
											fileName,
											quote = F, sep = ",", row.names = F)
				}
	)
)
```

Now let's see that these files were actually created

```{r check-files}
cancerFiles <- list.files(HW_DIR, pattern = "^cDatTest-.*csv$", full.names = T)
print(basename(cancerFiles))
```

Next we use `ldply` to read all the files and form a dataframe by concatenating
the information in each file.

```{r batch-read-files}
cancerFilesData <- ldply(cancerFiles, function(x) {
	tmpData <- read.table(file.path(x), header = T, sep = ",", row.names = NULL)
	cancerLoc <- gsub(".*cDatTest-(.*).csv$", "\\1", x)
	cancerLoc <- gsub("_", " ", cancerLoc)
	tmpData <- tmpData %>% mutate(cancerLocation = cancerLoc)
})
cancerFilesData$cancerLocation <- as.factor(cancerFilesData$cancerLocation)
```

Note that we could have also used `lapply` instead of `ldply`, but the resulting
object would be a list holding all the dataframes, and we would have to call
`dplyr::rbind_all` on it in order to make a dataframe from it.  Using `ldply`
is a nice convenient way to do it easier - it takes a list (of files), creates
a dataframe from every item (the dataframe for a specific cancer type), and
combines them all together into a dataframe. That's the plyr way!  
Let's just make sure our new dataset that we read matches the old one

```{r cancer-files-equal-test}
all.equal(cancerData, cancerFilesData)
```

Awesome. Now to get rid of this huge mess, let's delete all these files

```{r remove-cancer-files}
file.remove(cancerFiles)
```

### Data manipulation and reshaping
Suppose I don't care about the specific cancer types, but I just want to know
how many US citizens had or died from cancer each year.  I will show two
equivalent ways to approach this: using `plyr::ddply` and `dplyr::group_by`.
I personally perfer the `ddply` way, but I'm not sure if it's because it's
objectively more elegant or if it's because I'm more used to it.

```{r cancer-group-by-year}
cancersByYearPlyr <- ddply(cancerData,
													 .(year, stat),
													 summarize,
													 freq = sum(freq))
cancersByYearDplyr <- cancerData %>%
	group_by(year, stat) %>%
	summarize(freq = sum(freq)) %>%
	ungroup
```

Let's make sure that both methods resulted in the same dataframe

```{r check-group-by-equality}
identical(data.frame(cancersByYearPlyr), data.frame(cancersByYearDplyr))
```

We should also see what they actually look like to ensure it's what we wanted

```{r print-grouped-data}
print(head(cancersByYearDplyr))
```

Yup, looks good. 
Now I want to calculate the mortality rate from cancer in each year. I can't
figure out a nice way to do it within tidy data, so my solution is to reshape
the data to be wide in order to perform the calculation.  I also add a
_mortalityRateDrop_ variable that shows how much mortality rate changed from
the previous year, which isn't terribly useful but I just wanted to use the
`dplyr::lag` function.

```{r add-mortalityRate}
cancersByYear <- cancersByYearPlyr
cancersByYear <- tbl_df(cancersByYear) %>%
	spread(stat, freq) %>%
	mutate(mortalityRate = deaths/cases,
				 mortalityRateDrop = lag(mortalityRate) - mortalityRate)
print(cancersByYear)
```

### Add another source of data
One other thing I wanted to do is see how much of the population is affected
relative to the population size. I will read the file containing country populations
at different years, and extract from it only the US data for the relevant years,
and tidy it up a bit.  This example shows some more ways to be fancy with
`tidyr::gather`, `dplyr::mutate` and `dplyr::filter`

```{r read-pop-data}
popData <- tbl_df(read.csv(file.path(DATA_DIR, "worldPopByYear.csv")))
print(popData)
yearMin <- min(cancerData$year)
yearMax <- max(cancerData$year)
popDataClean <- popData %>%
	filter(Country.Code == "USA") %>%
	gather(year, population, starts_with("X")) %>%
	dplyr::select(year, population) %>%
	mutate(year = extract_numeric(year)) %>%
	filter(year %in% yearMin:yearMax)
print(popDataClean)
```

### Calculate deaths per million
Next we add the population of the US to every row, and now we can calculate
cancer deaths per million people. I'll remove the _population_ variable
after using it because it won't be needed any more

```{r calc-deathsPerM}
cancersByYear <- cancersByYear %>%
	left_join(popDataClean, by = "year") %>%
	mutate(deathsPerM = deaths / (population/1000000)) %>%
	dplyr::select(-population)
print(cancersByYear)
```

Very nice (well, we're looking at numbers of deaths... so I guess "nice"
isn't the right word to describe what we're seeing. But data-wise, very nice!)  
Now if we want to plot this information, it will be much easier to do so if we
get it back into a tidy tall/long format. So let's do that.

```{r tidy-big-data}
cancersByYear <- cancersByYear %>%
	gather(stat, value, -year) %>%
	arrange(year)
print(head(cancersByYear))
```

Alright, we're ready to plot. In this tidy format, it's very trivial to plot
all the pieces of information for a given year. I will be omitting the 
mortalityRateDrop variable since it's not that informative and plotting
4 variables looks much better than 5 (a 2x2 square vs ... awkwardness) 

```{r plot-big-data, fig.width = 11, fig.align = "center"}
p <-
	ggplot(cancersByYear %>% filter(stat != "mortalityRateDrop")) +
	geom_point(aes(x = as.factor(year), y = value), size = 3) + 
	geom_line(aes(x = as.factor(year), y = value, group = 1)) + 
	facet_wrap(~ stat, scales = "free_y") +
	theme_bw(20) +
	theme(axis.text.x = element_text(angle = 270, vjust = 0.5)) +
	xlab("year")
print(p)
ggsave(file.path(HW_DIR, "cancerCombinedStats.pdf"), p)
```

So pretty! And we're done!

---------------------
### Aside: Trick to reorder levels of factor based on dataframe row order
If you have a dataframe arranged in a specific way and you want the levels
of a factor to be ordered in the same order as the rows are, use
`df <- mutate(df, col = factor(col, col))`  
Example:

```{r }
df <- data.frame(
	num = 5:1,
	word = c("five", "four", "three", "two", "one"))
levels(df$word)
```

Levels are alphabetical by default

```{r }
df$word
```

Right now the order (of the dataframe) is 5,4,3,2,1

```{r }
df <- df %>%
	arrange(num) %>%   # rearrange the df in the order we want (1,2,3,4,5)
	mutate(word = factor(word, word)) # this line reorders the factor in the same order
levels(df$word)
```

Now the levels are 1,2,3,4,5
